Some examples in extended logic counting classes

### 2017-01-10 LICS/counting.tex
sc

Fix a signature $\R = \{R_1, \ldots, R_k\}$ and assume that $\A$ is an ordered (finite) $\R$-structure with a domain $A = \{a_1, \ldots, a_n\}$.
%$ containing $n$ elements. 
Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$. 
%(that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). 
Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the 
%following binary string~\cite{L04}:
%\begin{eqnarray*}
	%\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
%\end{eqnarray*}
$\enc(\A) =  0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A)$~\cite{L04}. 
%We define the class of all $\R$-functions, denoted by $\Func(\R)$, as the class of all functions $f: \ostr \rightarrow \bbN$.
%Given a function complexity class $\CC$ (i.e. $f: \Sigma^* \rightarrow \bbN$ for every $f \in \CC$), we say that a function $f \in \Func(\R)$ can be computed in $\CC$ if there exists $g \in \CC$ such that $f(\A) = g(\enc(\A))$ for every $\A \in \ostr$. 
%Note that the function $g$ outputs $f$ for encodings of structures and can behave arbitrarily otherwise.
We can now formalize the notion of capturing a counting complexity class. From now on, we assume that every counting complexity class $\CC$ consists of functions $f : \{0,1\}^* \to \N$ that are invariant under isomorphism of structures. More precisely, given a signature $\R$, a function $f : \{0,1\}^* \to \N$ is invariant under $\R$-isomorphisms if for every pair $\A, \mathfrak{B} \in \ostr[\R]$ such that $\A$ and 
$\mathfrak{B}$ are isomorphic, it holds that $f(\enc(\A)) = f(\enc(\mathfrak{B}))$. It is worth noticing that this is a usual assumption in descriptive complexity~\cite{L04}.
%over $\R$-structures.
\begin{definition} \label{def:cap}
	Let $\FF$ be a fragment of $\qso$ and $\CC$ a counting complexity class. Then {\em  $\FF$ captures $\CC$ over ordered $\R$-structures} if the  following conditions hold:
	\begin{enumerate}
		\item for every $\alpha \in \FF$, there exists $f \in \CC$ such that $\sem{\alpha}(\A) = f(\enc(\A))$ for every $\A \in \ostr[\R]$.

@@

What counting complexity classes can be captured with fragments of $\qso$?
For answering this question, it is reasonable to start with $\shp$, a well-known and widely-studied counting complexity class~\cite{arora2009computational}. 
Since $\shp$ has a strong similarity with $\np$, one could expect a ``Fagin-like'' Theorem~\cite{fagin1974generalized} for this class. 
Actually, in~\cite{SalujaST95} it was shown that the class $\sfo$ captures $\shp$.
In our setting, the class $\sfo$ is contained in $\eqso(\fo)$, which also captures $\shp$ as expected. 
\begin{proposition} \label{prop:capture-shP}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{proposition}
Recall  that every function class $\# \LL$ is contained in $\eqso(\LL)$ (see Section~\ref{sec:previous}). Thus, it directly follows from~\cite{SalujaST95}  that every $\shp$-function can be defined in $\eqso(\LL)$. The other direction of Proposition \ref{prop:capture-shP} follows by the fact that $\shp$ is closed under first- and second-order sum.
%Moreover, these closure properties also hold if we change the boolean core $\fo$ to any polynomial-time logic like $\lfp$. This gives the following corollary.
%\begin{corollary} \label{prop:capture-shP-LFP}
%	$\eqso(\lfp)$ captures $\shp$ over ordered structures.
%\end{corollary}

By following the same approach as~\cite{SalujaST95}, Compton and Gr\"adel~\cite{ComptonG96} show that $\seso$ captures $\spp$, where $\eso$ is the existential fragment of $\so$. As one could expect, if we parametrize $\eqso$ with $\eso$, we can also capture~$\spp$.
\begin{proposition} \label{prop:capture-spanP}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{proposition}
Can we capture $\fp$ by using $\# \LL$ for some fragment $\LL$ of $\so$? A first attempt could be based on the use of a fragment $\LL$ of $\so$ that capture either $\ptime$ or $\nlog$~\cite{G92}. Such an approach fails as $\# \LL$ can encode $\shp$-complete problems in both cases; in the first case, one can encode the problem of counting the number of satisfying assignments of a Horn  propositional formula, while in the second case one can encode the problem of counting the number of satisfying assignments of a 2-CNF propositional formula. A second attempt could be based then on considering a fragment $\LL$ of $\fo$. 
But even if we consider the existential fragment $\Sigma_1$ of $\fo$ the approach fails, as $\# \Sigma_1$ can encode $\shp$-complete problems like counting the number of satisfying assignments of a 3-DNF propositional formula\cite{SalujaST95}. One last attempt could be based on disallowing the use of second-order free variables in $\sfo$. But in this case one 
%The previous result connecting fragments from $\# \LL$ with $\shp$ and $\spp$ is the further that we can go by using SO-formulas with free variables. 
%So, what happen with other counting complexity classes like $\fp$ and $\fpspace$?
%By trying to search for an $\# \LL$ class for capturing $\fp$ one can easily see that this is pointless.
%Indeed, 
%By using free second-order variables one can easily capture $\shp$-hard functions (see Section~\ref{sec:syntactic}) and 
%By only using free first-order variables one 
cannot capture exponential functions definable in $\fp$ such as~$2^n$.
%(e.g. $2^{n}$), so
Hence, it is not  clear how to capture
%impossible 
%to capture 
$\fp$ 
%or $\fpspace$ 
by following the approach proposed in~\cite{SalujaST95}. 
%if we follow previous approaches like
On the other hand, if we consider our framework and move out from $\eqso$, we have other options for counting like first- and second-order products. In fact, the combination of $\qfo$ with $\lfp$ is exactly what we need to capture $\fp$.
% the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{theo:capture-fp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To prove this theorem, 
%capture $\fp$ 
one first shows that every formula in $\qfo(\lfp)$ can be evaluated in polynomial time. 
Indeed, $\lfp$ is a polynomial-time logic~\cite{I86,vardi1982complexity}, and the sum and product quantifiers can also be computed in polynomial time. 
For the other direction, one has to use $\qfo(\lfp)$ to simulate the run of a polynomial time TM $M$ computing a function, in particular using the quantitative quantifiers to reconstruct the natural number returned by $M$ in the output tape. 
%built in the ouptape. 
It is important to notice that the alternation between sum and products quantifiers is crucial for this reconstructions and, thus, crucial for capturing $\fp$.
%That is, without these operators is difficult to capture $\fp$.

@@

The results of this section validate $\qso$ as an appropriate logical framework for extending the theory of Descriptive Complexity to counting complexity classes. In the following sections, we provide more arguments for this claim, by considering some fragments of $\eqso$ and, moreover, by showing how to go beyond $\eqso$ to capture other 
%counting complexity 
classes.

>>>>

Fix a signature $\R = \{R_1, \ldots, R_k\}$ and assume that $\A$ is an ordered (finite) $\R$-structure with a domain $A = \{a_1, \ldots, a_n\}$.
%$ containing $n$ elements. 
Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$. 
%(that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). 
Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the string
%following binary string~\cite{L04}:
%\begin{eqnarray*}
	%\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
%\end{eqnarray*}
$0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A)$~\cite{L04}. 
%We define the class of all $\R$-functions, denoted by $\Func(\R)$, as the class of all functions $f: \ostr \rightarrow \bbN$.
%Given a function complexity class $\CC$ (i.e. $f: \Sigma^* \rightarrow \bbN$ for every $f \in \CC$), we say that a function $f \in \Func(\R)$ can be computed in $\CC$ if there exists $g \in \CC$ such that $f(\A) = g(\enc(\A))$ for every $\A \in \ostr$. 
%Note that the function $g$ outputs $f$ for encodings of structures and can behave arbitrarily otherwise.
We can now formalize the notion of capturing a counting complexity class. From now on, we assume that every counting complexity class $\CC$ consists of functions $f : \{0,1\}^* \to \N$ that are invariant under isomorphism of structures. More precisely, given a signature $\R$, a function $f : \{0,1\}^* \to \N$ is invariant under $\R$-isomorphisms if for every pair $\A, \mathfrak{B} \in \ostr[\R]$ such that $\A$ and 
$\mathfrak{B}$ are isomorphic, it holds that $f(\enc(\A)) = f(\enc(\mathfrak{B}))$. It is worth noticing that this is a usual assumption in descriptive complexity~\cite{L04}.
%over $\R$-structures.
\begin{definition} \label{def:cap}
	Let $\FF$ be a fragment of $\qso$ and $\CC$ a counting complexity class. Then {\em  $\FF$ captures $\CC$ over ordered $\R$-structures} if the  following conditions hold:
	\begin{enumerate}
		\item for every $\alpha \in \FF$, there exists $f \in \CC$ such that $\sem{\alpha}(\A) = f(\enc(\A))$ for every $\A \in \ostr[\R]$.

@@

What counting complexity classes can be captured with fragments of $\qso$?
For answering this question, it is reasonable to start with $\shp$, a well-known and widely-studied counting complexity class~\cite{arora2009computational}. 
Since $\shp$ has a strong similarity with $\np$, one could expect a ``Fagin-like'' Theorem~\cite{F75} for this class. 
Actually, in~\cite{SalujaST95} it was shown that the class $\sfo$ captures $\shp$.
In our setting, the class $\sfo$ is contained in $\eqso(\fo)$, which also captures $\shp$ as expected. 
\begin{proposition} \label{prop:capture-shP}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{proposition}
Recall  that every function class $\# \LL$ is contained in $\eqso(\LL)$ (see Section~\ref{sec:previous}). Thus, it directly follows from~\cite{SalujaST95}  that every $\shp$-function can be defined in $\eqso(\LL)$. The other direction of Proposition \ref{prop:capture-shP} follows by the fact that $\shp$ is closed under first- and second-order sum.
%Moreover, these closure properties also hold if we change the boolean core $\fo$ to any polynomial-time logic like $\lfp$. This gives the following corollary.
%\begin{corollary} \label{prop:capture-shP-LFP}
%	$\eqso(\lfp)$ captures $\shp$ over ordered structures.
%\end{corollary}

By following the same approach as~\cite{SalujaST95}, Compton and Gr\"adel~\cite{ComptonG96} show that $\seso$ captures $\spp$, where $\eso$ is the existential fragment of $\so$. As one could expect, if we parametrize $\eqso$ with $\eso$, we can also capture~$\spp$.
\begin{proposition} \label{prop:capture-spanP}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{proposition}
Can we capture $\fp$ by using $\# \LL$ for some fragment $\LL$ of $\so$? A first attempt could be based on the use of a fragment $\LL$ of $\so$ that capture either $\ptime$ or $\nlog$~\cite{G92}. Such an approach fails as $\# \LL$ can encode $\shp$-complete problems in both cases; in the first case, one can encode the problem of counting the number of satisfying assignments of a Horn  propositional formula, while in the second case one can encode the problem of counting the number of satisfying assignments of a 2-CNF propositional formula. A second attempt could be based then on considering a fragment $\LL$ of $\fo$. 
But even if we consider the existential fragment $\Sigma_1$ of $\fo$ the approach fails, as $\# \Sigma_1$ can encode $\shp$-complete problems like counting the number of satisfying assignments of a 3-DNF propositional formula\cite{SalujaST95}. One last attempt could be based on disallowing the use of second-order free variables in $\sfo$. But in this case one 
%The previous result connecting fragments from $\# \LL$ with $\shp$ and $\spp$ is the further that we can go by using SO-formulas with free variables. 
%So, what happen with other counting complexity classes like $\fp$ and $\fpspace$?
%By trying to search for an $\# \LL$ class for capturing $\fp$ one can easily see that this is pointless.
%Indeed, 
%By using free second-order variables one can easily capture $\shp$-hard functions (see Section~\ref{sec:syntactic}) and 
%By only using free first-order variables one 
cannot capture exponential functions definable in $\fp$ such as~$2^n$.
%(e.g. $2^{n}$), so
Thus, it is not  clear how to capture
%impossible 
%to capture 
$\fp$ 
%or $\fpspace$ 
by following the approach proposed in~\cite{SalujaST95}. 
%if we follow previous approaches like
On the other hand, if we consider our framework and move out from $\eqso$, we have other options for counting like first- and second-order products. In fact, the combination of $\qfo$ with $\lfp$ is exactly what we need to capture $\fp$.
% the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{theo:capture-fp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To prove this theorem, 
%capture $\fp$ 
one first shows that every formula in $\qfo(\lfp)$ can be evaluated in polynomial time. 
Indeed, $\lfp$ is a polynomial-time logic~\cite{I86,vardi1982complexity}, and the sum and product quantifiers can also be computed in polynomial time. 
For the other direction, one has to use $\qfo(\lfp)$ to simulate the run of a polynomial time TM $M$ computing a function, in particular using the quantitative quantifiers to reconstruct the natural number returned by $M$ in the output tape. 
%built in the ouptape. 
It is important to notice that the alternation between sum and product quantifiers is crucial for this reconstructions and, thus, crucial for capturing $\fp$.
%That is, without these operators is difficult to capture $\fp$.

@@

The results of this section validate $\qso$ as an appropriate logical framework for extending the theory of descriptive complexity to counting complexity classes. In the following sections, we provide more arguments for this claim, by considering some fragments of $\eqso$ and, moreover, by showing how to go beyond $\eqso$ to capture other 
%counting complexity 
classes.

### 2017-01-10 LICS/extentions.tex
sc

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{FFK94} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
%such that for some $\np$ machine, $f(x)$ is the number of accepting runs minus the number of rejecting runs.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
%it can compute negative values by definition. 
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to %describe functions
capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Similar than for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{example}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. Then this can be easily done in $\qsoz$ with the formula:
	$
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1 
	$.
\end{example}
Adding negative constants is a mild extension to allow subtraction in the logic. 
It follows from our characterization of $\shp$ that this is exactly what we need to capture  $\gp$.
\begin{corollary} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{corollary}
%The previous result is straightforward to prove from the characterization of $\shp$ and the definition of $\gp$. 
%However, 
%We 
%believe 
%think
This is an interesting result that shows how robust is $\qso$ when capturing different complexity classes.

A different class of functions comes from considering the optimization version of a decision problem. For example, one can define MAX-SAT as the problem of determining the maximum number of clauses, of a given CNF propositional formula, that can be made true by an assignment. Here, MAX-SAT is defined in terms of a maximization problem which in its essence differs from the functions in $\shp$. 
To formalize this set of optimization problems, Krentel defined $\optp$~\cite{krentel1988complexity} as the class of functions computable by taking the maximum or minimum of the output values over all runs of a polynomial-time NTM machine with output tape (i.e. each run produces a binary string which is interpreted as a number). 
For instance, MAX-SAT is in $\optp$ as several others optimization versions of $\np$-problems.
Given that in~\cite{krentel1988complexity} the author does not make the distinction between $\max$ or $\min$, in~\cite{vollmer1995complexity} the authors define the classes $\maxp$ and $\minp$ as the max and min version of the problems in $\optp$ (i.e. $\optp = \maxp \cup \minp$).

In order to capture classes of optimization functions, we can extend $\qso$ with $\max$ and $\min$ quantifiers (called $\optqso$) as follows. 
Given a signature $\R$, the set of $\optqso$-formulae over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{multline*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \\ \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{multline*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing, respectively, instead of computing a sum. 
\begin{example}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the largest clique in a graph. This can be done in $\optqso$ as follows:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
	Notice that formula $\sa{z} X(z)$ is used to compute the number of nodes in a clique $X$. 
\end{example}
Similar than for $\maxp$ and $\minp$, we have to distinguished between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulae constructed from $\qfo$ operators and $\max$-formulae $\max\{\alpha,\alpha\}$, $\maxa{x} \alpha$ and  $\maxa{X} \alpha$.
%operators of first- and second-order. 
The class $\minqso$ is defined analogously changing $\max$ with $\min$. Notice that in $\maxqso$ and $\minqso$, second-order sum and product are not allowed. For instance, formula $\alpha_6$ in Example \ref{ex:optqso} is in $\maxqso$.
%the formulae $\alpha_1$ and $\alpha_6$ are in $\maxqso$ but $\alpha_2$ and $\alpha_3$ are not. 
As one could expect, $\maxqso$ and $\minqso$ are the needed logics to capture $\maxp$ and $\minp$.
\begin{theorem} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ captures $\maxp$ and $\minp$, respectively, over ordered structures.
\end{theorem}
It is important to mention that a similar result was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ (resp., $\minpb$) of problems in $\maxp$ (resp., $\minp$) whose output value is polynomially bounded.
Interestingly, Theorem \ref{theo:capture-optp} is stronger since our logic has the freedom to use sum and product quantifiers, instead of using a max-and-count problem over Boolean formulae. 
Finally, it is easy to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the product $\Pi x$ in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.

>>>>

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{FFK94} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
%such that for some $\np$ machine, $f(x)$ is the number of accepting runs minus the number of rejecting runs.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
%it can compute negative values by definition. 
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to %describe functions
capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Similar than for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{example}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. This can be easily done in $\qsoz$ with the formula:
	$
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1 
	$.
\end{example}
Adding negative constants is a mild extension to allow subtraction in the logic. 
It follows from our characterization of $\shp$ that this is exactly what we need to capture  $\gp$.
\begin{corollary} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{corollary}
%The previous result is straightforward to prove from the characterization of $\shp$ and the definition of $\gp$. 
%However, 
%We 
%believe 
%think
This is an interesting result that shows how robust is $\qso$ when capturing different complexity classes.

A different class of functions comes from considering the optimization version of a decision problem. For example, one can define MAX-SAT as the problem of determining the maximum number of clauses, of a given CNF propositional formula, that can be made true by an assignment. Here, MAX-SAT is defined in terms of a maximization problem which in its essence differs from the functions in $\shp$. 
To formalize this set of optimization problems, Krentel defined $\optp$~\cite{krentel1988complexity} as the class of functions computable by taking the maximum or minimum of the output values over all runs of a polynomial-time NTM machine with output tape (i.e. each run produces a binary string which is interpreted as a number). 
For instance, MAX-SAT is in $\optp$ as several others optimization versions of $\np$-problems.
Given that in~\cite{krentel1988complexity} the author does not make the distinction between $\max$ and $\min$, in~\cite{vollmer1995complexity} the authors define the classes $\maxp$ and $\minp$ as the max and min version of the problems in $\optp$ (i.e. $\optp = \maxp \cup \minp$).

In order to capture classes of optimization functions, we can extend $\qso$ with $\max$ and $\min$ quantifiers (called $\optqso$) as follows. 
Given a signature $\R$, the set of $\optqso$-formulae over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{multline*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \\ \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{multline*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing, respectively, instead of computing a sum. 
\begin{example}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the largest clique in a graph. This can be done in $\optqso$ as follows:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
	Notice that formula $\sa{z} X(z)$ is used to compute the number of nodes in a set $X$. 
\end{example}
Similar than for $\maxp$ and $\minp$, we have to distinguished between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulae constructed from $\qfo$ operators and $\max$-formulae $\max\{\alpha,\alpha\}$, $\maxa{x} \alpha$ and  $\maxa{X} \alpha$.
%operators of first- and second-order. 
The class $\minqso$ is defined analogously changing $\max$ with $\min$. Notice that in $\maxqso$ and $\minqso$, second-order sum and product are not allowed. For instance, formula $\alpha_6$ in Example \ref{ex:optqso} is in $\maxqso$.
%the formulae $\alpha_1$ and $\alpha_6$ are in $\maxqso$ but $\alpha_2$ and $\alpha_3$ are not. 
As one could expect, $\maxqso$ and $\minqso$ are the needed logics to capture $\maxp$ and $\minp$.
\begin{theorem} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ capture $\maxp$ and $\minp$, respectively, over ordered structures.
\end{theorem}
It is important to mention that a similar result was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ (resp., $\minpb$) of problems in $\maxp$ (resp., $\minp$) whose output value is polynomially bounded.
Interestingly, Theorem \ref{theo:capture-optp} is stronger since our logic has the freedom to use sum and product quantifiers, instead of using a max-and-count problem over Boolean formulae. 
Finally, it is easy to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the product $\Pi x$ in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.
